### 2019-10-29 ms/ms.tex
Ready for bioRxiv

% To-do before bioRxiv:
% * p-values in lambda
% * change years in simulation
% * check taxa numbers in violin plot
% * add copyright information for phylopics
% * add sigma into the range simulation code
% * not a track index. But what is it?...
% * can you get more birds into the manuscript? They seem to be being dropped somewhere (perhaps at the pre-processing step?)
\section*{Title page}
\textbf{Article title}: The realised velocity of climate change
reveals remarkable idiosyncrasy of species' distributional shifts

@@

$^*$ To whom correspondence should be addressed:
\url{will.pearse@usu.edu} and \url{j.davies@ubc.ca}

\textbf{Word-count}: XXX (excluding figure legends)

\clearpage
\textbf{To date, our understanding of how species have shifted in
  response to recent climate warming has been based on a few studies
  with limited number of species. Here we present a comprehensive,
  global overview of species' distributional responses to changing
  climate across a broad variety of taxa (animals, plants, and
  fungi). We characterise species' responses using a metric that
  describes the realised velocity of climate change: how closely
  species' responses have tracked changing climate through time
  climate. In contrast to existing `climate velocity' metrics that have focused on space,
  we focus on species and their null expectation of change
  % The null expectation bit is new; I think this is a novelty, no?
  , and thus can examine drivers of inter-specific variation. Here we
  show that species are tracking climate on average, but not
  sufficiently to keep up with the pace of climate change. Further,
  species responses are highly idiosyncratic, and thus previous
  projections assuming uniform responses may be misleading. This is in
  stark contrast to species' present-day and historical niche
  dimensions, which show strong evidence of the imprint of
  evolutionary history and functional traits. Our analyses are a first
  step in exploring the vast wealth of empirical data on species'
  historic responses to recent climate change.}

@@

The main direct human-induced drivers that impact biodiversity now are
habitat loss \supercite{Newbold2015} and fragmentation
\supercite{Haddad2015}, but climate change is likely to become a
dominant driver in the next few
decades\supercite{Parmesan2003,Thuiller2007,Urban2015}. Projected
impacts of climate change on biodiversity have attracted much
attention, but the uncertainty around the magnitude of future
extinctions (\emph{e.g.}, 37\%\supercite{Thomas2004} vs.\
7.9\%\supercite{Urban2015} of species) means more work is needed.  In
the face of changing climate species must either shift in space, to
track favourable conditions, or time, for example, flowering and
breeding earlier, or migrating sooner. Evidence suggests they are
doing both \supercite{Parmesan2003,Parmesan2006,Menzel2006,Chen2011},
but not all species are responding equally. Some species are expanding
their distributions, sometimes invasively \supercite{Hellmann2008},
and some species are flowering later\supercite{Cook2012}; species may
also be simply becoming less predictable
\supercite{Pearse2017phenology}, perhaps as evolved responses to
environmental cues break down.  Simply quantifying species' mean
responses ignores the important variability in responses among species
\supercite{Wolkovich2012}. Developing robust predictions of ecological
responses to climate change thus requires data from diverse species,
ecosystems, and climates to account for this large inter-specific
variation.
% Better, I think, thank you. I think there was something about the
% previous phrasing that sounded a bit too much like we were knocking
% people's stats, rather than the concepts, if that makes sense?

The velocity of
temperature change---the rate a species would have to shift its range
to maintain constant temperature---has been calculated at 0.42 km/yr
\supercite{Loarie2009}. A meta-analysis of taxonomic groups estimated
the average pole-ward shift in species' distributions has been
approximately 1.69 km/yr \supercite{Chen2011},
suggesting that species might be capable of keeping pace with shifting
climate.
% You questioned the 1.69 figure - this text is from you! :D
% - I agree, though, that I'm not sure this estimate is fair. I looked
%   in the supplement and there's a remarkable lack of data going into
%   the estimate, to be honest with you... Is this sufficiently polite
%   to address that? They also don't address the file-drawer effect at all.
Yet this estimate is derived from just 21 studies with taxonomic bias
reflecting the state of the field at the time (361 birds, 9 mammals,
and no plants)
% I want to make sure we're nice about this point, because it's a
% meta-analysis so it's hardly their fault! Also one of my PhD
% supervisors is on this paper :p
, and critically not address the large variation in response among
species or within taxa. Here we leverage the vast volume of
distributional data stored in digital data
repositories\supercite{url_gbif} to provide the first global synthesis
of species distributional responses to climate change. By contrasting
historical niche dimensions with those of the present day, we can
characterise the realised biotic velocity of climate change, which
represents the actual velocity with which species track climate across
space.

We show that, on average, species' historical (pre-1980) ranges have
warmed significantly, and that species have generally shifted their
distributions to ameliorate this warming, but not sufficiently to keep
pace with climate change. Yet this general pattern Yet this general
pattern masks a remarkably degree of variability in species responses:
many species' ranges are now even warmer than if their distibutions
had remained static. Further, we find that, while species' niche
dimensions are strongly predicted by their evolutionary history and
functional traits, the same is not true of their degree of
tracking. Thus while species' distributions are reasonably
predictable, their changes in distributions are not.

\clearpage
\subsection*{Results and discussion}
\textbf{A comparative index of species' niche tracking.} We propose to
measure the extent to which a species' distribution is tracking
climate change by standardising its observed change by the change it
would have experienced had it not shifted its distribution. Formally,
we define this as:

@@

Where $current_\tau$ is a given quantile ($\tau$) of a species'
climatic distribution across its present-day range, $past_\tau$ the
same quantile of a species' range in the past, and $projected_\tau$
the same quantile of a species' present-day range under past climatic
conditions. This index should be calculated separately for
temperature, pollution, or whatever environmental variable is of
interest. Thus a $track$ of 0 would indicate a species that has
tracked changing climate perfectly, while a $track$ of 1 would be
consistent with a species that had not moved whatsoever to track
changing environmental conditions. Values between 0 and 1 suggest
moderate tracking, while values greater than 1 or less than 0
represent a change in climate relationship greater than if the species
had not moved at all. Uncertainty in either species' past or current
ranges can be accounted for through a bootstrap procedure, whereby the
past and/or current distributions are re-sampled (with replacement) to
match the sampling of the other time-period, and the standard
deviation of the index across those samples reported (see methods).

Our approach has two main innovations: its focus on (1) the whole of a
species' distribution and (2) expected change. (1) Species' range modelers typically focus on species'
limits\supercite{Sexton2009} (\emph{i.e.}, the upper and lower
quantiles) while community ecologists \supercite{Tilman2004} and
macro-evolutionary \supercite{Zanne2018} biologists focus on central
tendencies (\emph{i.e.}, the median). We see no strong
\emph{a priori} reason that species can, or must, track all quantiles
equally. Indeed some evidence suggests different constraints on species
warm range limts versus their cool range limits, with biotic interactions more influential in the former,
and climate determining the latter\supercite{Freeman2018}.
It is possible, therefore, that shifts at species' range edges may be disconnected
from changes towards the centre of their distribution.
% phrase better?
Our approach is to check all quantiles of a species' distribution. (2)
The denominator of our index re-scales species' change around an
expectation under distributional stasis. We suggest that species'
distributional responses should be measured on the basis of the
magnitude of change to which they are exposed. This change reflects
the magnitude of ecological and evolutionary selection, and while
there are many ways of addressing this, we suggest ours is perhaps the
simplest and most transparent, so perhaps represents a fair starting
point. We note that we could also rescale our index according to the
distance that would need to be travelled to maintain climate stasis
(\emph{sensu} \citeauthor{Loarie2009}\supercite{Loarie2009}), but
prefer to present here a dimensionless (and so unitless) index of
tracking, which is simpler to compare across taxa and between
biogeographic regions.
% You ask if we could make a map of this. We could, but I don't think
% we could by Tuesday...

\textbf{Niche tracking across the tree of life throughout the past
  sixty years.} To assess the extent to which species have tracked
climate, we used the most comprehensively sampled time-series data for
species' distributions (GBIF\supercite{url_gbif}) and climate (UEA's
CRU\supercite{Harris2014}). This represents over 320 Gb of occurrence
data, collected over the past century.
% Perhaps now you see why I keep having panic attacks about running
% everything through!...
We focus on five quantiles: the \nth{5},
\nth{25}, \nth{50} (median), \nth{75}, and \nth{95}. Here, we present species' range data collected from 1955--2015 for 10,700 species (4,879 plants,
1,517 fungi, 273 mammals, 433 birds, 147 reptiles and amphibians, and
3,451 insects), each observation matched to one of nine climate
variables in the year in which they were sampled (mean, minimum,
maximum temperature, diurnal temperature range, precipitation, 'rainy-day'
frequency, vapour pressure, and cloud cover). In the methods we
give more detail of our reproducible analysis pipeline, and in the
supplementary materials we repeat our analyses with different data
curation and cleaning choices. Our metric was designed to account for known biases in the GBIF data \supercite{Beck2014}, and
we present in the supplementary materials additional analyses that
address and account for such biases.

Figure \ref{violin} shows observed index values for mean temperature.
Almost all species, 99.5\%, are tracking some climate variable through time, and a large number of
species are, to some extent, tracking temperature through time (65\%
track at least one quantile, 43\% track at least two, and 4\% track
all five). The three temperature indices are the least-tracked of any
climate variable, while precipitation is the most-tracked (86\% of
species track at least one of its quantiles). Notably, however, many
taxa are not tracking climate through time, and many have experienced
more change than if their distributions had remained static (99.9\% of
species have `overshot' in at least three quantiles). As we show in
figure \ref{violin}, and explore further in the supplementary
materials, these results are consistent with our null simulations
where species' range change is a combination of essentially random
movement and a moderate degree of subsequent climate filtering.
% Jonathan is getting confused about the numbers
% 99% vs. XXX%; be clearer

\textbf{Idiosyncratic drivers of species' tracking.} To examine the
potential drivers of species' ranges and range tracking, we first
examined the phylogenetic signal of species' climate quantiles. We
focused on mammals\supercite{Faurby2015}, birds\supercite{Jetz2012},
amphibians \supercite{Jetz2018}, reptiles \supercite{Zheng2016}, and
plants\supercite{Smith2018}, since we were able to find broadly
inclusive, dated phylogenies for these taxa. We found statistically
significant evidence of phylogenetic signal across 158 of 180
taxon-climate-quantile combinations of species' past ranges (and 153
in the present; see figure \ref{phylogeny}). This is consistent with
the known patterns of phylogenetic conservatism outlined above. For
our index, however, we found evidence of phylogenetic signal in only
29 combinations. This confirms the known results that species'
current-day and past ranges are strongly influenced by evolutionary,
but suggests that the magnitude of change (relative to expectations)
may be driven by different drivers.
% Will: this is based on lambda *values* and so you need to update
% this with p-values...

We additionally examined species' traits as potential drivers of
species' distributions and potential to track changing environment. We
focused on mammals\supercite{Jones2009a,Wilman2014},
birds\supercite{Wilman2014}, and plants\supercite{Wright2004}, since
global, openly-released trait datasets exist for these taxa. Again, we
found statistically significant associations between traits and past
distributions of 193 of 240 taxon-climate-quantile combinations (203
in the present at $\alpha_{5\%}$; see figure \ref{traits}), but not
for between the traits and our tracking index (9 out of 240l
4\%). While the traits we used are strongly associated with species'
climate responses\supercite{Wright2004} (hence their strong predictive
power of species' distributions, but not tracking index), we emphasise
that we do not examine traits related to dispersal ability. This is
due to a paucity of open-access dispersal traits
\supercite{Gallagher2019}, and we highlight that future work should
address whether our index is explained by such drivers.
% Another potential problem here is no trait*env interaction, which I
% sense is also important...
% WILL - re-do TEMP.INDEX ~ TEMP! THERMOPHILES!!!

\textbf{Climate space is not the same as tracking space.} Given our
index is poorly predicted by traditionally important variables, we now
consider whether index values are predictive of each other. It is well
known that global climate\supercite{xxx}, and so species' positions
across the various axes of climate space\supercite{xxx}, are not
independent, and so we set out to test whether species' tracking
indices are also independent. As we show in figure \ref{pca}, a
principal components analysis of species' current, past, and projected
climate distributions shows strong correlations across climate
variables. Yet the index does not: species' relative degree of
tracking is both of much higher dimension, and shows relatively less
correlation across climates axes. We argue, therefore, that while
species' positions in climate space show strong associations, species'
relative tracking of their position in climate space does not.

If true, this result would explain the lack of association between
species' traits and evolutionary history with their degree of climate
tracking. The ecological and evolutionary rules that determine
species' climate relationships are changing, and thus the factors that
once are strong predictors of past and current distributions are not
predictive of the differences between them. A likely

Yet there are alternative
explanations, of which we highlight three: (1) that we have missed
fundamentally important environmental axes, (2) biotic interactions,
and (3) that our index highlights random variation in
distribution. (1) We fully acknowledge that we ignore potentially
relevant environmental drivers such as geology (and so soil), habitat,
and all Anthropogenic factors other than climate. We are not alone in
this simplification\supercite{xxx}, but we detect strong evolutionary
pattern in the factors we do examine in the data underlying our
index. (2) Competition and facilitation among species are likely
strong drivers of species' ability to track climate\supercite{xxx}. We
acknowledge it is entirely possible that ignoring these factors is the
driver of our result and, if so, we urge for an acceleration of the
inclusion of such information into broad-scale analyses
\supercite{xxx}. (3) If species were perfectly tracking climate
(\emph{i.e.}, $current_\tau$ is equal to $past_\tau$ in the absence of
sampling error or some other stochastic process) then our index values
would essentially be random numbers. There are good empirical evidence
that this is not the case \supercite{xxx} but, as we show in the
supplementary materials, simulations suggest out values are
incompatible with this result.

\subsection*{Conclusion}
Our results are consistent with the processes underlying species'
responses to climate change being distinct from those that generated
their ranges in the past. That is not to say that we do not detect the
imprint of ecological and evolutionary history, or that those earlier
processes are of predictive power for species distributions, but
rather that to understand change we must consider new drivers. We do
not know what those drivers are, although we highlight that biotic
interactions may play a part. We also highlight that our data,
although comprehensive, are not taxonomically complete (we address
only XXX species) and are necessarily spatially ($0.5^\circ$ grid
cells) and temporally (yearly) coarse given the extent (global and 60
years) of our study. We are not the first to suggest that, for large
numbers of species, the extent of climate tracking is idiosyncratic
\supercite{xxx}. These results are, however, consistent with the
paleo-ecological record. During and after the last ice-age,
`no-analogue' communities generated previously-unseen species
assemblages \supercite{xxx}. Once previous mass extinction events were
underway, species' extinctions became unpredictable
\supercite{xxx}. Scientists have been warning for decades that we are
beginning a sixth mass extinction;
% Soule 1999 but find something earlier
perhaps we are seeing the end of its beginning.

This consensus of species' past and current ranges broadly tracking
the same underlying climate optima is complicated by paleo-ecological
studies that have revealed greater variability of species' historic
ranges in previous centuries and millennia than previously thought
\supercite{Williams2007,Veloz2012,Maguire2015}.

\clearpage
% Figures
\clearpage
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth]{../figures/violin-tmp-0-5.pdf}
  \end{center}
  \caption{\doublespacing \textbf{Most species are tracking global
      temperature, but there is pronounced variation among species and
      clades.} The vertical axis shows the \nth{50} quantile of our
    tracking index of annual mean temperature (boostrapped to account
    for variation in sampling; see methods). On the left, our seven
    major taxonomic groups plotted with the median and \nth{25} and
    \nth{50} quantiles of their distributions highlighted. At the
    far-right of the plot, output from our simulations of species that
    are moving but either imperfectly tracking, or not tracking at
    all, are shown. In the main text, we highlight that the majority
    of species are tracking global climate, but there is profound
    variation among species and clades. Fungi, insects, and reptiles
    are, broadly, tracking the \nth{50} quantile of mean temperature
    well, while other taxa (\emph{e.g.}, mammals and plants) are not.}
  \label{violin}
\end{figure}

\clearpage
\begin{landscape}
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=1.4\textwidth]{../figures/phylo-tmp-50.pdf}
  \end{center}
  \caption{\doublespacing \textbf{Despite strong phylogenetic signal
      in species' static distributions, species' changes in
      distributions are not phylogenetically patterned.} This is a
    single sample from the posterior distribution of plant phylogenies
    used for our analysis\supercite{Smith2018}. The two oldest
    branches have been shortened for clarity of presentation
    (emphasised with dashed lines). Its internal branches are
    ancestral state reconstructions, assuming Brownian motion, of
    species' median mean-temperature in the past. Its tips are
    coloured according to the track index of the same underlying data
    (compare with figure \ref{violin}). Both are coloured according to
    the legend at the bottom of the figure. While this is but a single
    plot (see also figure \ref{correls} and the supplementary
    materials), this general pattern of strong phylogenetic signal of
    climate, but not of degree of tracking, is true of almost all of
    our underlying data (see main text).}
  \label{pca}
\end{figure}
\end{landscape}

\clearpage
\begin{figure}[h!]
  \begin{center}
    \begin{subfigure}{.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/traits-50.pdf}
      \caption{Trait correlations}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/physig-5.pdf}
      \caption{Phylogenetic signal}
    \end{subfigure}
  \end{center}
  \caption{\doublespacing \textbf{Species' static climate variables
      show strong trait correlations and phylogenetic signal, but
      their relative degree of tracking does not.} In (a), we show the
    correlation between the median of each climate variable (rows)
    with each of our traits (columns); colours are given at the right
    of the figure. In each cell, the outer, larger box depicts the
    correlation between species' present-day climate and the trait,
    while the smaller, inner box the correlation for species' relative
    tracking (whose size is proportional to intensity of
    correlation). The 3 (of 54) track-index correlations that are
    statistically significant (at $\alpha_{5\%}$) are highlighted with
    crosses. In (b), we show the phylogenetic signal (Pagel's
    $\lambda$\supercite{Pagel1999}) of the median of each taxonomic
    group's (columns) median species' climate variables and relative
    tracking indices (rows). As with (a), in each cell the outer,
    larger box depicts the correlation between species' present-day
    climate and the trait, while the smaller, inner box the
    correlation for species' relative tracking (whose size is
    proportional to phylogenetic signal strength). The single
    track-index metric for which phylogenetic signal was stronger than
    the underlying data is highlighted with a cross.}
  \label{traits}
\end{figure}

>>>>

% To-do before bioRxiv:
% ! p-values in lambda

\section*{Title page}
\textbf{Article title}: The realised velocity of climate change
reveals remarkable idiosyncrasy of species' distributional shifts

@@

$^*$ To whom correspondence should be addressed:
\emph{will.pearse@usu.edu} and \emph{j.davies@ubc.ca}

\textbf{Word-count}: 2634 (excluding figure legends)

\clearpage
\textbf{To date, our understanding of how species have shifted in
  response to recent climate warming has been based on a few studies
  with a limited number of species. Here we present a comprehensive,
  global overview of species' distributional responses to changing
  climate across a broad variety of taxa (animals, plants, and
  fungi). We characterise species' responses using a metric that
  describes the realised velocity of climate change: how closely
  species' responses have tracked changing climate through time. In
  contrast to existing `climate velocity' metrics that have focused on
  space, we focus on species and index their responses to a null
  expectation of change in order to examine drivers of inter-specific
  variation. Here we show that species are tracking climate on
  average, but not sufficiently to keep up with the pace of climate
  change. Further, species responses are highly idiosyncratic, and
  thus highlight that projections assuming uniform responses may be
  misleading.  This is in stark contrast to species' present-day and
  historical climate niches, which show strong evidence of the imprint
  of evolutionary history and functional traits. Our analyses are a
  first step in exploring the vast wealth of empirical data on
  species' historic responses to recent climate change.}

@@

The main direct human-induced drivers that impact biodiversity now are
habitat loss \supercite{Newbold2015} and fragmentation
\supercite{Haddad2015}, but climate change is likely to become a
dominant driver in the next few
decades\supercite{Parmesan2003,Thuiller2007,Urban2015}. Projected
impacts of climate change on biodiversity have attracted much
attention, but the uncertainty around the magnitude of future
extinctions (\emph{e.g.}, 7.9\%\supercite{Urban2015} vs.\
37\%\supercite{Thomas2004} of species) highlights the inadequacy of
current forecasts.  In the face of changing climate species must
either shift in space, to track favourable conditions, or time, for
example, flowering and breeding earlier.
% I've dropped the migration reference here because that's shifting in
% both space and time, and so sort of ruins the either/or structure
Evidence suggests they are doing both
\supercite{Parmesan2003,Parmesan2006,Menzel2006,Chen2011}, but not all
species are responding equally. Some species are expanding their
distributions, sometimes invasively \supercite{Hellmann2008}, and some
species are flowering later\supercite{Cook2012}; species may also be
simply becoming less predictable \supercite{Pearse2017phenology},
perhaps as evolved responses to environmental cues break down.  Simply
quantifying species' mean responses ignores the important variability
in responses among species \supercite{Wolkovich2012}. Developing
robust predictions of ecological responses to climate change thus
requires data from diverse species, ecosystems, and climates to
account for this large inter-specific variation.

The velocity of temperature change---the rate a species would have to
shift its range to maintain constant temperature---has been calculated
at an average of 0.42 km/yr \supercite{Loarie2009}. A meta-analysis of
taxonomic groups estimated the average pole-ward shift in species'
distributions has been approximately 1.69 km/yr \supercite{Chen2011},
suggesting that species might be capable of keeping pace with shifting
climate. Yet even this estimate is derived from just 21 studies with
limited taxonomic coverage (361 birds, 9 mammals, and no plants), and
critically any single point estimate does not reflect the large
variation in responses among species or within taxa. Here we leverage
the vast volume of distributional data stored in digital data
repositories\supercite{url_gbif}, representing over 10,000 species of
plants, fungi, mammals, birds, reptiles, amphibians, and insects, to
provide the first global synthesis of species distributional responses
to climate change. By contrasting historical niche dimensions with
those of the present day, we can characterise the realised biotic
velocity of climate change, which represents the relative degree to
which species track climate across space.

We show that, on average, species' historical (pre-1980) ranges have
warmed significantly, and that species have generally shifted their
distributions to ameliorate this warming, but not sufficiently to keep
pace with climate change. Yet this general pattern masks a remarkably
degree of variability in species responses, and many species' ranges
are now even warmer than if their distributions had remained
static. Further, we find that, while species' niche dimensions are
strongly predicted by their evolutionary history and functional
traits, the same is not true of their degree of tracking. Thus while
species' distributions are reasonably predictable, their changes in
distributions are not.

\clearpage
\subsection*{Results and discussion}
\textbf{A comparative index of species' niche tracking.} We propose a
measure that captures the extent to which a species' distribution is
tracking climate by standardising its observed niche change by the
change it would have experienced had it not shifted its
distribution. Formally, we define this as:

@@

Where $current_\tau$ is a given quantile ($\tau$) of a species'
climatic distribution across its present-day range, $past_\tau$ the
same quantile of a species' range in the past, and $projected_\tau$
the same quantile of a species' past range under present-day climatic
conditions. This index should be calculated separately for
temperature, pollution, or whatever environmental variable is of
interest. Thus a $track$ of 0 would indicate a species that has
tracked changing climate perfectly, while a $track$ of 1 would be
consistent with a species that had not moved whatsoever to track
changing environmental conditions. Values between 0 and 1 suggest
moderate tracking, while values greater than 1 or less than 0
represent a change in climate relationship greater than if the species
had not moved at all. Uncertainty in either species' past or current
ranges can be accounted for through a bootstrap procedure, whereby the
past and current distributions are re-sampled to standardise sampling
across time periods (see methods).

Our approach has two main innovations: its focus on (1) the whole of a
species' distribution and (2) expected change. (1) When estimating
species' range distributions we typically focus on species'
limits\supercite{Sexton2009} (\emph{e.g.}, upper and lower quantiles)
while in community ecological \supercite{Tilman2004} and
macro-evolutionary \supercite{Zanne2018} studies we tend to focus on
central tendencies (\emph{e.g.}, the median). However, there is no
strong \emph{a priori} reason that species can, or must, track all
quantiles equally. Indeed some evidence suggests different constraints
on species' warm-range limits versus their cool-range limits, with
biotic interactions more influential in the former, and climate
determining the latter\supercite{Freeman2018}.  It is possible,
therefore, that the drivers of species' range edges and centres may be
diverging under climate change. Our approach is to check all quantiles
of a species' distribution. (2) The magnitude of experienced and
projected climate change shows substantial geographical variation
\supercite{Loarie2009}, and so species' responses might thus be
expected to also vary in space. We suggest that species'
distributional responses should be measured on the basis of the
magnitude of change to which they are exposed, and the denominator of
our index re-scales species' change around an expectation under
distributional stasis.  This change reflects the magnitude of
ecological and evolutionary selection, and while there are many ways
of addressing this, we suggest ours is perhaps the simplest and most
transparent, so represents a fair starting point. We note that we
could also rescale our index according to the distance that would need
to be travelled to maintain climate stasis (\emph{sensu}
\citeauthor{Loarie2009}\supercite{Loarie2009}), but prefer to present
here a unitless index of tracking, which is simpler to compare across
taxa and between biogeographic regions.

\textbf{Niche tracking across the tree of life throughout the past
  sixty years.} To assess the extent to which species have tracked
climate, we used the most comprehensively sampled time-series data for
species' distributions (GBIF\supercite{url_gbif}) and climate (UEA's
CRU\supercite{Harris2014}). This represents over 630 million raw
occurrence records collected over the past century. We focus on five
quantiles: the \nth{5}, \nth{25}, \nth{50} (median), \nth{75}, and
\nth{95}. Here, we present species' range data collected from
1955--2015 for 10,700 species (4,879 plants, 1,517 fungi, 273 mammals,
433 birds, 147 reptiles and amphibians, and 3,451 insects), each
observation matched to one of nine climate variables in the year in
which they were sampled (mean, minimum, maximum temperature,
precipitation, `rainy-day' frequency, vapour pressure, potential
evapotranspiration, `frost day' frequency, and cloud cover). In the
methods we give more detail of our reproducible analysis pipeline, and
in the supplementary materials provide instructions for repeating our
analyses with different data curation and cleaning choices. Our metric
was designed to account for known biases in the GBIF data
\supercite{Beck2014}, and in the methods we outline our simulations
which address the influence of sampling uncertainty.

Almost all species are tracking some climate variable through time
(99.5\% have at least one $track_\tau$ between 0 and 1). Most species
are, to some extent, tracking temperature (see Figure \ref{overview});
65\% track at least one quantile and 43\% track at least two.
% Giving an index in units of distance is a good idea; I think it
% needs some thought, though, because our index is unitless because we
% divide the units through; I'm not sure the number we get out of this
% would be directly interpretable like theirs.
The three temperature indices are, however, the least-tracked of any
climate variable, while precipitation is the most-tracked (86\% of
species track at least one of its quantiles). Notably, however, many
taxa are not tracking climate through time, and many have experienced
more change than if their distributions had remained static. Just as
most species are tracking at least one quantile of one climate
variable, 99.9\% of species have `overshoot' (have a $track_\tau$ less
than 0; for example becoming colder in a warming region) in at least
one quantile of at least one climate variable. As we show in Figure
\ref{overview}, these results are consistent with our null simulations
where species' range change is a combination of essentially random
movement and a moderate degree of subsequent climate filtering. Thus
while our results confirm that, by-and-large, species are tracking
climate change, the magnitude of this varies across species, taxa, and
aspects of climate (see Figure \ref{overview}). For example, while
fungi and insects are, on the whole, broadly tracking the median of
temperature (median $track_{50\%}$ of 0.46 and 0.64, respectively),
plants and mammals track temperature (median $track_{50\%}$ of 1.16
and 1.07, respectively) more poorly than precipitation (median
$track_{5\%}$ of 0.79 and 0.89, respectively).

\textbf{Idiosyncratic drivers of species' tracking.} To examine the
potential drivers of species' ranges and range tracking, we first
examined the phylogenetic signal of species' climate quantiles. We
focused on mammals, birds, reptiles, and plants, since we were able to
find broadly inclusive, dated phylogenies for these taxa. We found
strong evidence of phylogenetic signal in species' past (median
Pagel's $\lambda$\supercite{Pagel1999} of 0.60) and current (median
$\lambda$ of 0.57) distributions, but limited signal in our index
(median $\lambda$ of $<$0.001). Past climate showed stronger
phylogenetic signal than our index in 158 of all 180 (4 taxa, 5
quantiles, and 9 climate variables) possible taxon-climate-quantile
combinations (144 for current climate; see Figures \ref{phylo} and
\ref{traits-signal}). This confirms the known result that species'
current-day and past ranges are strongly influenced by
evolution\supercite{Wiens2010}, but suggests that species' realised
velocity of change (our tracking index) may be associated with
different factors.

We additionally examined species' traits as potential drivers of
species' distributions and potential to track changing environment. We
focused on mammals, birds, and plants, since global, openly-released
trait datasets exist for these taxa. Again, we found statistically
significant associations between traits and past distributions of 191
of 270 (at $\alpha_{5\%}$; 6 traits, 9 climate variables, and 5
quantiles) taxon-climate-quantile combinations (193 for current
distributions), but not between the traits and our tracking index (13
combinations or 5\%; see Figure \ref{traits-signal}). While the traits
we used are strongly associated with species' climate
responses\supercite{Wright2004,McCain2014} (hence their strong
predictive power of species' distributions, but not degree of
tracking), we emphasise that we could not find sufficient data to
assess the potential impact of dispersal
ability\supercite{Schloss2012}. These findings, while surprising, are
in keeping with emerging evidence that species' traits are a
relatively poor predictor of change at range edges
\supercite{Angert2011}.

\textbf{Species' movement through climate space is disconnected from
  their position within it.} Given our index is poorly predicted by
traditionally important variables, we now consider whether index
values are predictive of each other. It is well known that the axes of
global climate are not independent and that they are not changing
independently\supercite{Harris2014,IPCC2014}, and so we would expect
species' relative tracking of climate to show similar patterns. As we
show in Figure \ref{pca}, principal components analysis of species'
current and past climate distributions, which we refer to as their
climate space, shows strong correlations across climate variables. Yet
our index does not; species' relative degree of tracking is both of
much higher dimension (the amount of variance explained by each
principal component axis is similar) and shows relatively less
correlation across climates axes . We argue, therefore, that while
species' positions in climate space show strong associations, species'
relative tracking of their position in climate space does not. This
result explains the poor predictive power of species' traits and
evolutionary history for our index: the ecological and evolutionary
rules that determine species' climate relationships are
changing. Thus, in our analysis, we can explain species' current
distributions only by the degree to which they resemble past
distributions.

There are alternative explanations for our results, of which we
highlight three: (1) we are ignoring important niche axes, (2) we are
ignoring biotic interactions, and (3) our index highlights random
variation in distribution. (1) Our climate variables are not
all-encompassing, and we ignore other drivers such as geology and
habitat-type. Yet we do detect strong pattern in current and past
distributions, suggesting that the absence of pattern in our indices
is meaningful. (2) Competition and facilitation among species are
likely strong drivers of species' ability to track
climate\supercite{Davis1999}. Yet, again, they are surely also likely
to have driven species' past distributions, and so it is unclear to us
why only our metric, and not also current and past distributions, is
so poorly predicted by our trait and phylogenetic data. (3) If species
were perfectly tracking climate, $current_\tau$ would be equal to
$past_\tau$ and so our index values could reflect sampling error or
some other purely stochastic input. Yet such noise, if truly random,
would mean our empirical index values would be centred at 0 or
resemble our simulated data, neither or which is the case (see Figure
\ref{overview}).

\clearpage
\subsection*{Conclusion}
Our results are consistent with the processes underlying species'
responses to climate change being distinct from those that generated
their distributions in the past. This is in keeping with emerging
evidence from the paleo-ecological literature that species' historic
ranges were more variable than previously thought, and seemingly
influenced by different environmental factors than today
\supercite{Veloz2012,Maguire2015}. During and after the last ice-age,
for example, `no-analogue' conditions generated previously-unseen
species assemblages \supercite{Williams2007}. Equally, our observation
of idiosyncrasy in species' degree of tracking is consistent with
species' population crashes (and so extinctions) becoming less
predictable during previous mass extinction events
\supercite{Levinton2001,Jablonski2004}. We call upon others to examine
the additional drivers of our metric of relative change, and to extend
its definition to include more nuanced definitions of climate
space. We predict, however, that if the current mass extinction event
continues, it is likely that patterns of idiosyncrasy among species'
declines and distributional changes will become more common.

\clearpage
% Figures
\clearpage
\begin{figure}[h!]
  \begin{center}
    \begin{subfigure}{.7\textwidth}
      \includegraphics[width=\textwidth]{../figures/violin-tmp-0-5.pdf}
      \caption{$track_{50\%}$(mean-temp)}
    \end{subfigure}
    \\
    \begin{subfigure}{.7\textwidth}
      \includegraphics[width=\textwidth]{../figures/violin-pre-0-05.pdf}
      \caption{$track_{5\%}$(precipitation)}
    \end{subfigure}
  \end{center}
  \caption{\textbf{Most species are tracking some aspect of climate,
      but there is pronounced variation among species and clades.} In
    (a) we show the \nth{50} quantile (median) of our tracking index
    of mean annual temperature, and in (b) the \nth{5} quantile of our
    index of precipitation. On the left of each plot, our seven major
    taxonomic groups plotted (in green for plants and fungi, and in
    red for animals). At the far-right of (b), we show the output from
    our simulations of species that are perfectly tracking,
    imperfectly tracking, or not tracking climate at all (in blue; see
    Methods for details). To the right of each distribution of points
    the median (thickest line) and inter-quartile range (thinner
    lines) are shown as horizontal lines for each distribution. While
    the majority of species are tracking global climate to some
    extent, there is is profound variation among species and
    clades. Figure \ref{overview}a makes clear the variation among and
    within taxonomic groups; the distributions of fungi, insects, and
    reptiles are, broadly, tracking changes in mean temperature well,
    while other taxa (\emph{e.g.}, mammals and plants) are
    not. Comparing Figures \ref{overview}a and \ref{overview}b makes
    clear how different taxonomic groups and species may be tracking
    different aspects of climate; plants are broadly tracking the
    lower-limit of precipitation, for instance.}
  \label{overview}
\end{figure}

\clearpage
\begin{landscape}
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=1.4\textwidth]{../figures/phylo-tmp-50.pdf}
  \end{center}
  \caption{\textbf{Despite strong phylogenetic signal in species'
      static distributions, species' changes in distributions are not
      phylogenetically patterned.} Here we show the plant tree of
    life, with the two oldest branches shortened for compactness
    (emphasised with dashed lines). Internal branches are shaded by
    ancestral state reconstructions, assuming Brownian motion, of
    species' median mean-temperature in the past. Phylogeny tips are
    coloured according to the track index of the same underlying data
    (compare with Figure \ref{overview}). This general pattern of
    strong phylogenetic signal of climate (shown by the coloured
    pattern to the internal branches), but not of degree of tracking,
    is characteristic of almost all of our underlying climate data
    (see also Figure \ref{traits-signal}).}
  \label{phylo}
\end{figure}
\end{landscape}
